Writing Trash Bin

### 2024-01-04 paper.qmd
finished context section and moved bits that didnt fit to other sections

@@

# 2. Defining Uncertainty

## 2.1 Uncertainty is goal dependent
Statistics is, at its core, the study of uncertainty. Therefore discussing uncertainty visualisation a a separate sub domain to "normal" data visualisation is inherently confusing. What is typically meant by "uncertainty" visualisation is "noise", that is, we want to present some signal cushioned by its natural variance. Unfortunately, this distinction between "signal" and "noise" is entirely goal dependent.

The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of analysis. This conceptual way of thinking about uncertainty has been rattling around in the philosophical discussions of uncertainty for quite a while.

@Otsuka2023 points out that the philosophy of statistics is build upon taking real world entities and boiling them down into probabilistic objects and this ontological process is largely dependent on our goals. For example, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.

@Meng2014 shows that the entire concept of signal and noise are relative and the noise for one goal can be the signal for another. This break down is done by considering the data source, the "resolution" of the problem and data, and the phases involved in the analysis. @Meng2021 highlights that data needs to be treated as an output from another process rahter than an value neutral input.

@Carlin2023 highlights a range of pitfalls that come up frequently when statisticians use regression models and they connect these pitfalls to the teaching of statistical methods as goal agnostic tools. @Carlin2023 suggest that every research question belongs to one of three distinct types: descriptive, predictive, and causal, each of which has its own appropriate statistical methods.

@Amar2005

Since uncertainty visualisation is largely goal dependent, hard and fast uncertainty visualisation rules are difficult to design. Additionally communication of statistics cannot ignore the ontological process discussed by @Otsuka2023 because the goal of communication is to map those statistical objects back to real world entities.

After all, the distinction between uncertainty and signal is largely determined by the goal of an analysis rather than any inherent properties of our data.

This leads to several unanswered questions. Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models? Is it to supress signal that could be the result of random variance? Is it to find the best way to convey the mass function of a random variable? A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.

- Should the visualisations change based on these goals?

**need for task based understanding**
- several review papers acknowledge this issue is causing 
- **papers that talk about task specific stuff**
- This problem of a task based philosiphy extends all the way back to philisophical uncerstandings of statistics and variance.
- **notes from the Xiao-li Meng papers**
- **notes from john paper**
- While the statistics side brings up the purpose of a model or estimate for a philisophical consideration, it is a real, practical, and urgent problem for uncertainty visualisation. 
- **go through papers with endless different goals**
- **review paper that has an entire paragraph discussing differnet findings for box plots based on the task**

**need for statistics links**
- (no visual distinction between sample/pop variance) @Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision. (**also the paper sherry sent me**)

Uncertainty can only be defines within a specific problem. There is not a universal motivation agnostic "uncertainty" similarly as there is no motivation agnostic "signal". Therefore, in our following definitions of uncertainty, we will assume that the uncertainties we discuss are connected to some specific question.

## 2.2 Taxonomies
### 2.2.1 Taxonomies of Uncertainty
More commonly uncertainty is defined using a taxonomy rather than a strict definition.

@@

@@

In this section we will first go over the current state of uncertainty visualisation research, which is currently a collection of evaluation studies and review papers, and discuss what it is doing well and what could be done better. Following that, we will provide general reccomendations from the literature.

# 4 The current landscape of uncertainty visualisation research
# 4.1 The landscape
In section 2.1 it was mentioned that uncertainty must be defined within a specific motivating question, otherwise it inherently does not make sense. A large difficulty with the uncertainty visualisation evalusation studies is that this rule is not followed. There are a shockingly large number of evaluation studies that seem to pay no attention to the information that is relevant to the question they are asking, and show participants a selection of seemingly random visualisations from a statistical point of view.

@@

# 4.2 Key points from the literature

>>>>

- @Suh2023 found that the main issue with deployment of AI was communication between data science and subject matter experts, since subject matter experts didn't want technical presentations that didn't explain exactly when and how a model would fail.

@@

# 2. Defining Uncertainty
Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature. @Lipshitz1997 even commented that “there are almost as many definitions of uncertainty as there are treatments of the subject”. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.

## 2.1 Motivating Questions
Signal and noise can only be untangled in the presence of a motivating question. While it is seldom mentioned in the uncertainty visualisation literature, but is still important to understand that "uncertainty" does not exist without a motivating question. The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis. At the philosophical level, applied statistics is simply taking real world entities and boiling them down into probabilistic objects, an ontological process that is largely dependent on our goals [@Otsuka2023]. When we move onto data provenance the issue pursists, as what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the resolution of the question [@Meng2014]. After moving onto modelling this issue continues as each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to signal devoid of meaning [@Carlin2023]. Even at the final stages of visualisation a lack of understanding of the motivating question make it difficult to untangle what is signal and what is noise, leaving many uncertainty visualisation studies with conflicting results [@Kinkeldey2014]. These cases highlight that the signal that is being conveyed and the noise that hampens that signal cannot be untangled and therefore communicated without a predetermined motivating question. Therefore any discussion of uncertainty must be done within the context of a motivating question as it is exactly what makes the distinction between signal and noise.

## 2.2 Taxonomies
Within the context of a motivating uncertainty can be classified and 
### 2.2.1 Taxonomies of Uncertainty

More commonly uncertainty is defined using a taxonomy rather than a strict definition.

@@

*Notes from other section*
- The usually uncertainty is only considered at the final stage but it needs to be considered through all stages. The big data paradox shrinks confidence intervals but magnifies bias [@Bradley2021] because the confidence intervals shrink due to the sample size but the imprecise target group creates bias.
- @Meng2021 highlights that data needs to be treated as an output from another process rahter than an value neutral input.
- If the distinction between noise and uncertainty is largely task dependent, considerations of quantifying and expressing uncertainty should be considered for the entire duraction of a project; another concept that has been independently identified by several authors [@Kinkeldey2014; @Refsgaard2007]
- The most encompassing definition of uncertainty I have seen comes from @utypo who define uncertainty as **"any deviation from the unachievable ideal of completely deterministic knowledge of the relevant system"**. This definition encapsulates many concepts adjacent to randomness such as predictions, probability mass distributions (PMF), estimate error, and any data that is not a set of deterministic outcomes. This is the definition of uncertainty I will use for the rest of this report.
- While quantification of the uncertainty that arrises from statistical philosophy and data provenance are beyond the scope of this paper, how this uncertainty is communicated is not. The ontological process discussed by @Otsuka2023 where information is tossed as we boil our information down into statistical objects requires us to map our statistical objects back to real world entities. This means the noise introduced and signal lost in earlier stages of the data analysis need to be identified, quantified and expressed in terms of real world outcomes. Visualisation is often the final stage of larger data science project and the down stream effects of poor abstraction and problem at the lower levels become impossible to ignore [@Munzner2009].

@@

# 4 The current landscape of uncertainty visualisation research
In this section we will first go over the current state of uncertainty visualisation research, which is currently a collection of evaluation studies and review papers, and discuss what it is doing well and what could be done better. Following that, we will provide general reccomendations from the literature.
# 4.1 The landscape
In section 2.1 it was mentioned that uncertainty must be defined within a specific motivating question, otherwise it inherently does not make sense. A large difficulty with the uncertainty visualisation evalusation studies is that this rule is not followed. There are a shockingly large number of evaluation studies that seem to pay no attention to the information that is relevant to the question they are asking, and show participants a selection of seemingly random visualisations from a statistical point of view.

@@

Notes from previous sections that were moved here
- The concept of uncertainty being task dependent is *particularly* salient for uncertainty visualisation, and it is repeatedly identified as a problem in previous reviews of the uncertainty visualisation literature [@Kinkeldey2014; @Hullman2016] as well as across many sub domains and applications [@Wallsten1997; @Munzner2009; @Fischhoff2014; @Meng2021; @Amar2005]. The fact that this conclusion is repeatedly reached shows both the importance and the lack of acknowledgement this concept receives.
- @Fischhoff2014 discusses how decisions with different levels of flexibility should have different uncertainty visualisations. If the user has no choice they should simply be made aware of the risk, if they have several options, the user needs to see the uncertainty in each, if they need to come up with the options themselves, they need to see how things work and how the system fails.
- @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations.
- Given that uncertainty estimation and communication seem to be largely task depenent, what questions should we be testing? This leads to several unanswered questions. Is our goal in uncertainty communication to acknowledge and convey the inherrent randomness that exists in all statistical calculations and models? Is it to supress signal that could be the result of random variance? Is it to find the best way to convey the mass function of a random variable? A survey of the literature will leave the answers to these questions unclear, however this inherrent issue has not gone unnoticed.
- Statistics is, at its core, the study of uncertainty. Therefore discussing uncertainty visualisation a a separate sub domain to "normal" data visualisation is inherently confusing. What is typically meant by "uncertainty" visualisation is "noise", that is, we want to present some signal cushioned by its natural variance. Unfortunately, this distinction between "signal" and "noise" is entirely goal dependent.
- @Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision. (**also the paper sherry sent me**)
# 4.2 Key points from the literature

### 2024-01-04 writingdumpster.qmd
finished context section and moved bits that didnt fit to other sections

>>>>

# context section outtakes
This conceptual way of thinking about uncertainty has been rattling around in the different areas of data science for a while.
For example, when we record coin flips we typically want to model the behavior with a binomial distribution, so we ignore the outcome of a coin landing on its side despite the fact it is a real world possibility.
@Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters.  @Munzner2009 created a nested model for visualisation that highlighted how pitfalls in upstream levels, such as problem characterisation and data abstraction, can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it.
